---
title: "ViralQuantSim: rationale, model, and default calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ViralQuantSim: rationale, model, and default calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(viralquantsim)
```

## Motivation

Internal-control normalisation --- correcting a target measurement against a
reference signal measured alongside it --- is standard across quantitative
molecular assays. Examples include housekeeping-gene normalisation and the
comparative-Ct (Delta Ct) method in qPCR, and spike-in controls in sequencing
workflows. Its benefit, however, is conditional: a reference can only remove the
technical variation it *shares* with the target, and in exchange it contributes
its own measurement noise. Whether normalisation helps therefore depends on the
balance of those two quantities.

Paired datasets in which the true target load is known, alongside target and
internal-control measurements, are rarely public. ViralQuantSim does not
reconstruct any specific study. Instead it makes the data-generating assumptions
explicit and asks a methodological question: under which technical conditions
does an internal control improve, leave unchanged, or worsen quantitative
estimation?

## The core idea: shared versus specific variation

Every technical effect in the simulation is split into a **shared** component
(applied identically to the target and the internal control) and a
**specific** component (drawn independently for each). Delta Ct is the
difference between target and internal-control Ct. Because the shared component
enters both channels equally, it cancels in the subtraction; the specific
components do not. Delta Ct is therefore blind to shared noise but exposed to
specific noise.

This yields a variance trade. Forming Delta Ct removes the shared variance (a
gain) but injects the internal control's own specific variance (a cost).
Normalisation improves accuracy --- lower root-mean-square error against the
true load --- only when the shared variance removed exceeds the
control-specific variance introduced.

## The generative model

`simulate_assay()` draws a known true load, applies shared and analyte-specific
recovery, reverse-transcription, and inhibition effects, then produces qPCR Ct
values (with technical replicates and detection-limit censoring) and
sequencing-derived read counts (overdispersed, with detection-limit dropout).

```{r}
params <- default_parameters()
params$n_samples <- 300
simulated <- simulate_assay(params)
head(simulated[, c("true_log10_viral_load", "target_ct", "delta_ct", "log10_target_reads")])
```

## Candidate models

Five interpretable linear models estimate the true load from measurable
quantities: target Ct; Delta Ct; log10 reads; reads plus target Ct; and reads
plus Delta Ct. The comparisons of interest are target Ct versus Delta Ct (does
the control help qPCR?) and reads versus reads plus Delta Ct (does the control
add to a sequencing-based estimate?).

```{r}
models <- fit_candidate_models(simulated)
compare_models(models, simulated, cross_validate = FALSE)[, c("model", "rmse", "r_squared")]
```

## Default calibration

The default parameters are not fitted to any dataset. They describe a generic
quantitative-sequencing regime and each choice is a statement about assay
physics rather than about any specific study:

- **Library noise** (`library_shared_sd`, `target_library_specific_sd` = 0.10):
  a well-optimised preparation contributes modest per-sample variation.
- **Overdispersion** (`sequencing_concentration` = 400): a deduplicated count
  assay sits close to Poisson sampling.
- **Detection limit** (`dropout_midpoint_log10` = 2.0, `dropout_steepness` =
  4.0): a limit of detection near 100 copies/mL --- a standard clinical
  threshold --- with a sharp cutoff.

Under these defaults, read count alone is a moderately strong quantitative
predictor (R-squared around 0.8) and dropout is confined to the low-abundance
tail. Because the internal control cancels shared recovery and
reverse-transcription variance, which is present across the dynamic range, its
benefit is broad rather than tail-specific.

## Interpreting the sensitivity metric

`run_sensitivity_grid()` reports `delta_rmse`, the RMSE of the reads-plus-Delta
Ct model minus that of the reads-only model. Negative values mean the
internal-control-adjusted model performed better; values near zero mean it added
little; positive values mean adjustment worsened prediction.

```{r}
grid <- run_sensitivity_grid(
  "shared_recovery_sd", c(0.1, 0.5),
  "ic_specific_recovery_sd", c(0.05, 0.4),
  repetitions = 2L
)
aggregate(delta_rmse ~ value_x + value_y, data = grid, FUN = mean)
```

As expected from the variance trade, the adjustment helps most when shared
recovery variation is large and control-specific variation is small, and helps
least (or hurts) in the opposite regime.

## Scope

ViralQuantSim is a simulation and benchmarking tool. It is not a clinical
viral-load calculator, a validated diagnostic device, or a substitute for
wet-laboratory validation.

## References

Livak KJ, Schmittgen TD (2001). Analysis of relative gene expression data using
real-time quantitative PCR and the 2^-Delta Delta CT method. *Methods*, 25(4),
402--408.
